---
title: "p8105_hw3_zx2482"
author: "Allison Xia"
date: "2023-10-11"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(ggplot2)

library(p8105.datasets)
```
## Problem 1


```{r}
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```

#### Answer questions about the data

This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart |> select(product_id) |> distinct() |> count()` products found in `r instacart |> select(user_id, order_id) |> distinct() |> count()` orders from `r instacart |> select(user_id) |> distinct() |> count()` distinct users.

```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

In total, there are 134 aisles, and fresh vegetables and fresh fruits are the most items ordered by far.

```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```


```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```
The three most popular items in aisles `baking ingredients`, `dog food care`, and `packaged vegetables fruits`, and includes the number of times each item is ordered in the table above.

```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```
Pink Lady Apples are generally purchased slightly earlier in the day than Coffee Ice Cream, with the exception of day 5.

## Problem 2

```{r Load data}
data("brfss_smart2010") 

brfss_smart2010 = 
  brfss_smart2010 |> 
  as_tibble()
```

```{r clean brfss_smart2010}
brfss_overall_health =
  brfss_smart2010 |> 
  rename(state = Locationabbr,
         location = Locationdesc) |> 
  janitor::clean_names() |> 
  filter(topic == "Overall Health") |> 
  filter(response  %in% c("Excellent","Very good","Good","Fair","Poor")) |> 
  arrange(desc(response)) |> 
  mutate(response = ordered(response, levels = c("Poor", "Fair", "Good","Very good","Excellent"))) |> 
  arrange(response) 
```

#### In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r}
states_2002 <- 
  brfss_overall_health |> 
  filter(year == "2002") |> 
  group_by(state) |> 
  summarize(number_locations= n_distinct(location)) |> 
  filter(number_locations >= 7) |> 
  pull(state)

states_2010 <- 
  brfss_overall_health |> 
  filter(year == "2010") |> 
  group_by(state) |> 
    summarize(number_locations = n_distinct(location)) |> 
  filter(number_locations >= 7) |> 
  pull(state)
```

In 2002, these following states were observed at 7 or more locations: `r states_2002`. In 2010, these following states were observed at 7 or more locations: `r states_2010`

#### Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a "spaghetti" plot of this average value over time within a state (that is, make a plot showing a line for each state across years -- the geom_line geometry and group aesthetic will help).

```{r Create brfss_excellent}
brfss_excellent = 
  brfss_overall_health |> 
  filter(response == "Excellent") |> 
  group_by(state, year) |> 
  summarize (mean_data_value = mean(data_value, na.rm=TRUE))
```

```{r Plot brfss_excellent}
ggplot(brfss_excellent, aes(x=year, y=mean_data_value, color=state)) +
  geom_line() +
  labs(title="Data average value over years within a state",
       x="Year",
       y="Mean Data Value")
```

This linegraph shows the pattern of mean BRFSS data value for all states from 2002 to 2010. This plot is too busy to reveal much information about the data, indicating that we need to perform further exploratory data analysis (EDA).

#### Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses ("Poor" to "Excellent") among locations in NY State.

```{r}
brfss_NY_responses = 
  brfss_overall_health |> 
  filter(state == "NY" & (year =="2006" | year == "2010"))
```

```{r}
ggplot(brfss_NY_responses, aes(x= response, y = data_value))+
  geom_boxplot(fill="skyblue", color="black")+
  labs(title="Boxplot showing distribution of data value for responses in NY", x="Responses", y="Data Value") +
  facet_wrap(~year)
```

This boxplot shows distributions of BRFSS data `data_value` for responses in NY in 2006 and 2010.The median `data_value` for responses as `Poor` in 2006 is `r brfss_NY_responses |> filter(response == "Poor" & year == 2006) |> summarize(median(data_value))`. The median `data_value` for responses as `Fair` in 2006 is `r brfss_NY_responses |> filter(response == "Fair" & year == 2006) |> summarize(median(data_value))`. The median `data_value` for responses as `Good` in 2006 is `r brfss_NY_responses |> filter(response == "Good" & year == 2006) |> summarize(median(data_value))`. The median `data_value` for responses as `Very good` in 2006 is `r brfss_NY_responses |> filter(response == "Very good" & year == 2006) |> summarize(median(data_value))`. The median `data_value` for responses as `Excellent` in 2006 is `r brfss_NY_responses |> filter(response == "Excellent" & year == 2006) |> summarize(median(data_value))`. The median `data_value` for responses as `Poor` in 2010 is `r brfss_NY_responses |> filter(response == "Poor" & year == 2010) |> summarize(median(data_value))`. The median `data_value` for responses as `Fair` in 2010 is `r brfss_NY_responses |> filter(response == "Fair" & year == 2010) |> summarize(median(data_value))`. The median `data_value` for responses as `Good` in 2010 is `r brfss_NY_responses |> filter(response == "Good" & year == 2010) |> summarize(median(data_value))`. The median `data_value` for responses as `Very good` in 2010 is `r brfss_NY_responses |> filter(response == "Very good" & year == 2010) |> summarize(median(data_value))`. The median `data_value` for responses as `Excellent` in 2010 is `r brfss_NY_responses |> filter(response == "Excellent" & year == 2010) |> summarize(median(data_value))`. The distribution for data value is right-skewed for `Poor`,`Very good` responses in 2006, and `Very good`,`Good` responses in 2010. The distribution for data value in other responses are left-skewed.

## Problem 3

#### Load, tidy, merge, and otherwise organize the data sets

```{r Load NHANES datasets}
accelerometer = read_csv("./accelerometer.csv")
demographic = read_csv("./demographic.csv",skip =4)
```

```{r Clean demographic}
demographic_tidy =
  demographic |> 
  janitor::clean_names() |> 
  na.omit() |> 
  filter(age >= 21) |> 
  mutate(sex = factor(sex),
         education = factor(education))
```

```{r Clean and tidy accelerometer}
accelerometer_tidy=
  accelerometer |> 
  janitor::clean_names() |> 
  pivot_longer(
    min1:min1440, 
    names_to = "min",
    names_prefix = "min",
    values_to = "mims"
  ) |> 
  mutate(min = as.numeric(min))
```

```{r Merge two datasest}
nhanes = demographic_tidy |> 
  left_join(accelerometer_tidy,by = join_by(seqn))
```

#### Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items

```{r}
sex_education = 
  demographic_tidy |> 
  group_by(sex, education) |>
  summarize(count = n())|>
  mutate(sex = recode(sex, "1" = "Male", "2" = "Female"),
         education = recode(education, "1" = "Less than High School", "2" = "High school equivalent", "3"="More than high school"))
sex_education |> 
  pivot_wider(
    names_from = sex,
    values_from = count
  ) |> 
  knitr::kable(digits = 2) 
```
* Both males and females have the highest count in the "More than high school" education level. For males, the number of participants increases with each increase in education level. For females, there are more participants with "Less than High School" education compared to those with a "High school equivalent". In the "More than high school" category, females (`r sex_education |> filter(sex == "Female" & education == "More than high school") |>pull(count)`) slightly outnumber males `r sex_education |> filter(sex == "Male" & education == "More than high school") |>pull(count)`). In contrast, males (`r sex_education |> filter(sex == "Male" & education == "High school equivalent") |>pull(count)`) have more participants than females (`sex_education |> filter(sex == "Female" & education == "High school equivalent") |>pull(count)`) in the "High school equivalent" category.

```{r Visualization}
my_labels<- c("1" = "Less than High School", "2" = "High school equivalent", "3"="More than high school")

ggplot(demographic_tidy, aes(x = sex, y = age)) +
  geom_boxplot(fill="skyblue", color="black") +
  facet_wrap(~education,labeller = labeller(.default = my_labels))+
  labs(title="Boxplot of age distribution for women and men in each education category", x=" Sex (1 = Male, 2 = Female) ", y="Age")
```

This boxplot shows age distribution for men and women in each eduation category.The median age for participants with an education `Less Than High School` is `r demographic_tidy |> filter(sex == 1 & education == 1) |> summarize(median(age))` for men, and `r demographic_tidy |> filter(sex == 2 & education == 1) |> summarize(median(age))` for women. Both distributions in `Less Than High School` category are right-skewed. The median age for participants with an education `High school equivalent` is `r demographic_tidy |> filter(sex == 1 & education == 2) |> summarize(median(age))` for men, and `r demographic_tidy |> filter(sex == 2 & education == 2) |> summarize(median(age))` for women. Both distributions in `High school equivalent` category are right-skewed.The median age for participants with an education `More than high school` is `r demographic_tidy |> filter(sex == 1 & education == 3) |> summarize(median(age)) `for men, and `r demographic_tidy |> filter(sex == 2 & education == 3) |> summarize(median(age))` for women. Both distributions in `More than high school` category are left-skewed.

#### Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r Data Wrangling}
nhanes_total_activities = 
  nhanes |> 
  group_by(seqn) |> 
  mutate(total_activity = sum(mims)) |> 
  select(seqn, age, education, sex, total_activity) |> 
  distinct()
```

```{r Plot}
ggplot(nhanes_total_activities, aes(x = age, y = total_activity, color = sex))+
  geom_point()+
  facet_wrap(~education,labeller = labeller(.default = my_labels))+
  geom_smooth(method="lm", se=FALSE)+
  labs(title="Scatterplot for correlation between age and total activities", x="Age", y="Total Activities", color = "Sex (1 = Male, 2 = Female)")
```
* This scatterplot shows the relationship between total activities and age, bifurcated by gender, and further categorized into three tabs based on educational attainment: 'Less than High School', 'High School Equivalent', and 'More than High School'. 
* A discernible trend across all categories and genders indicates a negative correlation between total activities and age. Participants with an education 'Less than High School' start with highest activities when young but shows the steepest decline in total activities as they age when compared to  other education categories. Participants with an education `Equivalent to High School` show the second steepest decline and those with an education `More than High School` show the most gradual decline in activities with age. 
* The slope of the decline in total activities with age is similar for both males and females.In the 'High School Equivalent' and 'More than High School' categories, females demonstrate higher total activities across all ages compared to their male counterparts. In the 'Less than High School' category, total activities for males and females are relatively the same across the age spectrum.

#### Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

```{r}
nhanes |> 
  group_by(min, education,sex) |> 
  arrange(min) |> 
  summarize(mims_median = median(mims))|>
  ggplot(aes(x = min, y = mims_median, color = sex)) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm, formula = y ~ x, se = FALSE) +
  labs(title = "Linegraph for Activity Over the Course of the Day",
       x = "Time in Minute",
       y = "MIMS Value (Activity)",
       color = "Sex (1 = Male, 2 = Female)") +
  facet_wrap(~education,labeller = labeller(.default = my_labels))
```

The scatterplot demonstrates the pattern in the activity levels over the course of the day of all participants. The activity initially increases till the noon and subsequently decreases. For participants in the 'less than high school' category, males exhibit a steeper slope in activity level over the course of the day compared to females. Conversely, in the other two categories, females display a more pronounced increase in activity level over the course of the day, as indicated by steeper slopes.
